Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-07-17 surfaces-write-up/surface-write-up.tex
small edit to last lemma

\begin{example}
\label{eg:hyperplane}
Suppose $D = \frac{1}{2}H_0 - \frac{1}{3}H_1$ where $H_0 = V(x_0),
H_1 = V(x_1)$ are two coordinate hyperplanes in $\bp^2$. Then, $R_D$
has generators in degree $2$ and $3$ which can be written as $u^2
\frac{x_1}{x_0}, u^3 \frac{x_1}{x_0}.$ In fact, for all degrees
less than $5$ the elements of $R_D$ can all be expressed as
rational functions in $x_0, x_1$. However, in degree $6$, there is
the rational function $\frac{x_1^2 x_2}{x_0^3}$. Since this
involves $x_2$, it must be a generator.

This example generalizes slightly to any divisor of the form $D =
\frac{1}{k}H_0 - \frac{1}{k+1}H_1 \in \di \bp^2 \otimes \bq,$ with
$k \in \bn$, showing that there will always exist a generator in
degree $a(a + 1)$.

This example further generalizes to the following situation:
Suppose $D = \sum_{i=1}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m
\otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm
_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$
where all $f_i$ can be written as a polynomial function in $x_0,
\ldots, x_{m-1},$ we have $R_D$ always has a generator in degree
$\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of
~\cite[Theorem 8]{dorney:canonical} of generation in degree at most
$\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is
that the divisors may ``be contained in a hyperplane,'' i.e.~ be
expressible as functions in $m$ of the $m+1$ variables on $\bp^m$.
This problem can easily be circumvented by adding in ``ghost
points.'' That is, we may add divisors of the form $0 \cdot H_i$ to
$D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$,
then $f_0, \ldots, f_n$.

@@

\begin{lem}
\label{lem:proj-relations-psi}
$\ker(\psi)$ is generated in degrees up to
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=1}^n \ell_i a_i.
\end{align}

@@

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying
$\bida_i + \sum_{j = 1}^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of
$f_i$ when viewed as a polynomial in $k[f_1,\ldots, f_{m + 1}]$.
Furthermore, the relations given in Equation
~\eqref{eqn:proj-relation-degree} generate all relations, since they
allow us to reduce any $ \prod_{j = 1}^n f_j^{r_i}$ to a canonical
form, with $r_i = -\lfloor  d \alpha_i\rfloor$ whenever $i  > m + 1$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. To
complete the proof, it suffices to bound the degree of generation
of the relations of the form $(z_i - \beta_i) \prod_{j = 1}^n
z_j^{c_j}$, by Equation ~\eqref{eqn:proj-relation-degree}. We first
associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1,
\ldots, c_n) \in \Sigma,$ as defined in Equation
~\eqref{eqn:Sigma-def}. Let $\Sigma_i \subseteq \mathbb{Z}^{n + 1}$
be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying
$r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.
Let
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)},
	\ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}

\noindent
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\br_{\geq 0}}
(e_1, \ldots, e_n)$ with $e_i$ as defined in Equation
~\ref{defn:e-i-proj}. Therefore, we can write any element of
$\Sigma_i$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

@@

\begin{thm}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \bp^m$. Let
$f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that
$D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]

\noindent
with relations generated in degree at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(	\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}

\begin{proof}
The bound on degree of generation is precisely the content of Lemma 
~\ref{lem:proj-generators}. It only remains to bound the degree of 
relations.

By ~\ref{lem:bound-ker-chi}, $\ker(\chi)$ is generated in degrees 
up to $2\sum_{i=1}^n \ell_i a_i$ and by Lemma
~\ref{lem:proj-relations-psi}, $\ker(\psi),$ is generated in
degrees up to $\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i$. 
Consequently, Lemma ~\ref{lem:composite-map} implies that $\ker(\phi)$
is generated in degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}
	(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{proof}

\section{Canonical Rings on Hirzebruch surfaces}
\label{sec:hirz}
In this section, the main aim is to prove Theorem
~\ref{thm:hirz-generators-relations}, bounding the degree of
generators and relations of the canonical ring of a $\bq$-divisor
on any Hirzebruch surface.

@@

By Proposition ~\ref{prop:cone-generation}, in order to determining a spanning set for $(R_D)_d,$ it suffices to find the extremal rays of $\Sigma.$
To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the sub-cone $\Sigma_1 \subset \Sigma$
given by
\begin{align*}
	\Sigma_1 = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d \alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i (\bida_i+\bidb_i) = 0 \}.	
\end{align*}
Then, $\Sigma_1$
has extremal rays given by

@@

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\[
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D \right)
\]

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]

\noindent
by
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

@@

\noindent
for all $d \in \mathbb{N}$ \todo{Aaron: This confuses me because $d$ is not used above, is $d$ supposed to be $i$?}
 and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{i=1}^n \bidb_j c_j = 0$.

>>>>

\begin{example}
\label{eg:hyperplane}
Suppose $D = \frac{1}{2}H_0 - \frac{1}{3}H_1$ where $H_0 = V(x_0), H_1 = V(x_1)$ are two coordinate hyperplanes in $\bp^2$. Then, $R_D$ has generators in degree $2$ and $3$ which can be written as $u^2 \frac{x_1}{x_0}, u^3 \frac{x_1}{x_0}.$ In fact, for all degrees less than $5$ the elements of $R_D$ can all be expressed as rational functions in $x_0, x_1$. However, in degree $6$, there is the rational function $\frac{x_1^2 x_2}{x_0^3}.$ Since this involves $x_2$, it must be a generator.

This example generalizes slightly to any divisor of the form $D = \frac{1}{k}H_0 - \frac{1}{k+1}H_1 \in \di \bp^2 \otimes \bq,$ with $k \in \bn$, showing that there will always exist a generator in degree $a(a+1).$

This example further generalizes to the following situation: Suppose $D = \sum_{i=1}^{n} \frac{\num_i}{\den_i}D_i \in \di \bp^m \otimes \bq,$ where $\deg D_i = \bida_i$ and $\deg D = \frac{1}{\lcm_{1 \leq i \leq n}(\den_i \cdot \bida_i)}$. Then, if $D_i = V(f_i)$ where all $f_i$ can be written as a polynomial function in $x_0, \ldots, x_{m-1},$ we have $R_D$ always has a generator in degree $\lcm(\den_i \cdot \bida_i)$.
\end{example}

\begin{rem}
\label{rem:ghost-motivation}
Example ~\ref{eg:hyperplane} shows that the naive generalization of ~\cite[Theorem 8]{dorney:canonical} of generation in degree at most $\sum_{i=1}^{n}\ell_i$ cannot possibly hold. The reason for this is that the divisors may ``be contained in a hyperplane,'' i.e., be expressable as functions in $m$ of the $m+1$ variables on $\bp^m$. This problem can easily be circumvented by adding in ``ghost points.'' That is, we may add divisors of the form $0 \cdot H_i$ to $D$, and reorder so that if $D = \sum_{i=1}^{n}\alpha_i V(f_i)$, then $f_0, \ldots, f_n$.

@@

\begin{lem}
\label{lem:proj-relations-psi}
$\ker(\psi)$ is generated in degrees up to
\begin{align}
\label{eqn:pm-relation-degree}
	\frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=1}^n \ell_i a_i.
\end{align}

@@

\noindent
for all $d \in \mathbb{N}$ and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$.

We define the $\beta_i$ above so that $\beta_i$ is the expansion of $f_i$ when viewed as a polynomial in $k[f_1,\ldots, f_{m+1}]$. Furthermore, the relations given in Equation ~\eqref{eqn:pm-relation-degree}
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{r_i}$ to a canonical form, with $r_i = -\lfloor d \alpha_i\rfloor$ whenever $i  > m + 1$.

For the remainder of the proof, fix $i \in \{1,\ldots, n\}$. 
To complete the proof, it suffices to bound the degree of generation of the relations of the form $(z_i - \beta_i) \prod_{j=1}^n z_j^{c_j}$, by 
Equation ~\eqref{eqn:pm-relation-degree}.  We first associate $(z_i - \beta_i)\prod_{j=1}^n z_j^{c_j}$ with $(d, c_1, \ldots, c_n) \in \Sigma,$ as defined in Equation ~\eqref{eqn:Sigma-def}.  
Let $\Sigma_i \subseteq \mathbb{Z}^{n+1}$ be the set of points of the form $(d, r_1, \ldots, r_n)$ satisfying $r_j \ge -d \alpha_j$ for all $j$ and $\sum_{j=1}^n r_j a_j = -a_i$.  Let 
\begin{align*}
	\delta_i = \left(\frac{a_i}{\deg(D)}, -\frac{\alpha_1 a_1}{\deg(D)}, \ldots, - \frac{\alpha_n a_n}{\deg(D)} \right).
\end{align*}
Then we see $\Sigma_i \subseteq \delta_i + \newspan_{\br_{\geq 0}}(e_1, \ldots, e_n)$ with $e_i$ as defined in Equation \ref{defn:e-i-proj}.
Therefore, we can write any element of $\Sigma_i$ as
\[
	\delta_i + \sum_{j=1}^n r_j e_j
\]
where $r_j \in \mathbb{R}$ for each $j$.

@@

\begin{thm}
\label{thm:proj-generators-relations}
Let $D = \sum_{i=1}^n \alpha_i D_i \in \bq \otimes \di \bp^m$. Let $f_i$ be homogeneous polynomials in $x_0, \ldots, x_m$ such that $D_i = V(f_i)$ for all $i$; suppose $\deg(f_i) = a_i$. 
Then $R_D$ is generated in degree at most 
\[
	\sum_{i=1}^n \ell_i a_i
\]
with relations generated in degree at most
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{thm}
\begin{proof}
The bound on degree of generation is precisely the content of Lemma ~\ref{lem:proj-generators}. It only remains to bound the degree of relations.
By ~\ref{lem:bound-ker-chi}, $\ker(\chi)$ is generated in degrees up to $2\sum_{i=1}^n \ell_i a_i$ and by Lemma ~\ref{lem:proj-relations-psi}, $\ker(\psi),$ is generated in degrees up to $\frac{\bida_i}{\deg(D)} + \sum_{i=1}^n \ell_i a_i.$ Consequently, Lemma ~\ref{lem:composite-map} implies that $\ker(\phi)$ is generated in degrees less than
\[
	\max \left(2 \sum_{i=1}^n \ell_i a_i, \frac{\max_{1\le i \le n}(\bida_i)}{\deg(D)} + \sum_{i=1}^n \ell_i a_i \right).
\]
\end{proof}

\section{Canonical Rings on Hirzebruch surfaces}
\label{sec:hirz}

In this section, the main aim is to prove Theorem ~\ref{thm:hirz-generators-relations}, bounding the degree of generators and relations of the canonical ring of a $\bq$-divisor on any hirzebruch surface.

@@

By Proposition ~\ref{prop:cone-generation}, in order to determining a spanning set for $(R_D)_d,$ it suffices to find the extremal rays of $\Sigma.$
To do this, we extend the method of O'Dorney \cite[Theorem 8]{dorney:canonical}. 
We first consider the subcone $\Sigma_1 \subset \Sigma$
given by
\begin{align*}
	\Sigma_1 = \{ (d, c_1, \ldots, c_n) \in \bz_{\geq 0}^{n+1} : c_i \ge - d \alpha_i \text{ for all }i \text{ and }\sum_{i=1}^n c_i (\bida_i+\bidb_i) = 0 \}.	
\end{align*}
Then, $\Sigma_1$
has extremal rays given by

@@

For $i \in \Tp(D)$ and $j \in \Tm(D)$, define
\[
	\epsilon_{i, j} := \left(\prod_{k \ne i, j} f_k^{-\alpha_k \ell_{i,j} (\bida_i \bidb_
	j - \bida_j \bidb_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) D \right)
\]

@@

Let $x_1, \ldots x_r$ be the generators of $R_D$ in degrees less than $\sigma$ (as given by Lemma ~\ref{lem:hirz-generators}), and let $\phi$ be the surjection $k[x_1, \ldots x_r] \to R_D$.  We can factor $\phi$ through the semigroup ring 
\[
	\bk[\Sigma] = \left \langle u^d z_1^{c_1} \cdots z_n^{c_n} : c_i \geq -d
	\alpha_i, \sum_{i=1}^{n} \bida_i c_i = \sum_{i=1}^{n} \bidb_i c_i
	\right \rangle.
\]
by
\[
\begin{tikzcd}
	\bk[x_1,\ldots, x_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
	x_i \ar r & u^{d_i}z_1^{c_{i1}} \cdots z_n^{c_{in}} \ar r & u^{d_i}f_1^{c_{i1}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]

@@

\noindent
for any degree $d \in \mathbb{N}$ \todo{Aaron: This confuses me because $d$ is not used above, is $d$ supposed to be $i$? Peter: Fixed?}
 and $c_i \ge -\alpha_i d$ satisfying $\bida_i + \sum_{j = 1}
^n \bida_j c_j = 0$ and $\bidb_i + \sum_{j=1}^n \bidb_j c_j = 0$.
